---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rfintext

<!-- badges: start -->
<!-- badges: end -->

Main purpose for this package is to:   
* Share text mining data sets easily.  
* Provide a few simple pre processing functions. Nothing special only wrappers 
around dplyr and other tidyverse functions.  
* Document all the steps for reproducible data analysis.  

> NOTE! It is strongly recommended to use functions directly from other 
packages since all the helper functions provided here are highly likely 
to change many times before I decide how they should look like and function.

This package contains following data sets:  
- `aspol`: Finnish housing policy documents.  
- `strategia`: Strategy documents of Finnish municipalities.    
Other data sets might be added later. Currently all the functionalities are 
tested on `aspol` data set only.

## Installation

You can install the development version of rfintext like so:

``` r
# install.packages("devtools")
devtools::install_github("StranMax/rfintext")
```


```{r setup, message=FALSE, warning=FALSE}
library(rfintext)
library(tidyverse)
```

## Data

Number of documents in `aspol` data set `r length(unique(aspol$kunta))`

Sentence nro 43 from Espoo document: 

```{r, R.options = list(width = 10000, pillar.print_max = 26, pillar.print_min = 26)}
aspol |> filter(kunta == "Espoo", sent == 43)
```

Top 20 words:  

```{r, R.options = list(width = 10000, pillar.print_max = 20, pillar.print_min = 20)}
aspol |>
  dplyr::filter(UPOSTAG %in% c("NOUN", "VERB", "ADJ", "ADV")) |>
  dplyr::count(LEMMA, UPOSTAG, sort = TRUE)
```

Power of lemmatization, number of different word forms with same lemma:  

```{r}
aspol |>
  group_by(LEMMA) |>
  summarise(taivutusmuodot = length(unique(FORM))) |>
  arrange(desc(taivutusmuodot))
```

10 most common word forms with lemma "asunto":  

```{r , R.options = list(width = 10000, pillar.print_max = 10, pillar.print_min = 10)}
aspol |>
  filter(LEMMA == "asunto") |>
  count(FORM, FEATS, sort = TRUE)
```

Word forms with most different classes:  

```{r}
aspol |>
  group_by(FORM) |>
  summarise(unique_sanaluokat = length(unique(UPOSTAG))) |>
  arrange(desc(unique_sanaluokat))
```

Different word forms, word classes and inflections for single lemma "olla":  

```{r , R.options = list(width = 10000, pillar.print_max = 20, pillar.print_min = 20)}
aspol |>
  filter(LEMMA == "olla") |>
  count(FORM, LEMMA, UPOSTAG, FEATS, sort = TRUE)
```

