---
title: "2 Document term matrix"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Converting tidytext format to numerical matrix form ready to fed different analysis.

```{r setup, message=FALSE, warning=FALSE}
library(rfintext)
# devtools::install_github("StranMax/rfinstats")
library(rfinstats)
library(quanteda)
library(tidytext)
library(stringr)
library(dplyr)
```

Turn word data frame into document term count data frame where each row is giving
term count in single document. Then we use `tidytext::bind_tf_idf()` to add also 
term frequency and term frequency weighted by document frequency.

```{r}
lemma <- aspol_filtered |>
  count(kunta, LEMMA, name = "dtf") |>
  bind_tf_idf(LEMMA, kunta, dtf)
```

Let's create variable `y` we are trying to explain with word frequencies.

```{r}
taantuvat <- taantuvat
y <- taantuvat[c("kunta", "suht_muutos_2010_2022", "luokka")] |>
  inner_join(lemma[c("kunta")]) |>
  group_by(kunta) |>
  slice(1)
```

We create data with only those docs we have y value. Some documents include multiple
municipalities and for now we leave those out.

```{r}
data <- left_join(lemma, y) |>
  filter(!is.na(LEMMA), !is.na(suht_muutos_2010_2022), !is.na(luokka)) |>
  select(kunta, LEMMA, dtf, tf, tf_idf, suht_muutos_2010_2022, luokka)

stopifnot(y$kunta == unique(data$kunta))
```

Three different document-term-matrices:

* n = with word counts  
* tf = with term frequencies  
*tf_idf = with term frequency weighted by document count [tf-idf](https://en.wikipedia.org/wiki/Tf%E2%80%93idf)

```{r}
tc <- cast_dfm(data = data, document = kunta, term = LEMMA, value = dtf)
docvars(tc, c("suht_muutos", "luokka")) <- c(y$suht_muutos_2010_2022, y$luokka)
tf <- cast_dfm(data = data, document = kunta, term = LEMMA, value = tf)
docvars(tf, c("suht_muutos", "luokka")) <- c(y$suht_muutos_2010_2022, y$luokka)
tf_idf <- cast_dfm(data = data, document = kunta, term = LEMMA, value = tf_idf)
docvars(tf_idf, c("suht_muutos", "luokka")) <- c(y$suht_muutos_2010_2022, y$luokka)
```


```{r}
# usethis::use_data(tc, overwrite = TRUE)
# usethis::use_data(tf, overwrite = TRUE)
# usethis::use_data(tf_idf, overwrite = TRUE)
```


We save 4 objects for playing with Python Scikit-learn package:

1. ID_labels (kunnat)  
2. Feature_names (LEMMA)  
3. Data (doc-term-matrix)  
4. y-variable (väkiluvun muutos)

```{r}
# save_data <- function(X, y, dir_out = tempdir()) {
#   
#   id_label <-  X |> as.matrix() |> row.names()
#   readr::write_lines(id_label, file.path(dir_out, "id_label.txt"), sep = ";")
#   
#   feature_name <- X |> as.matrix() |> colnames()
#   readr::write_lines(feature_name, file.path(dir_out, "feature_name.txt"), sep = ";")
#   
#   data <- X |> as.matrix() |> as.data.frame()
#   readr::write_csv(data, file.path(dir_out, "data.csv"))
#   
#   label <- y$suht_muutos_2010_2022
#   readr::write_lines(label, file.path(dir_out, "label.txt"), sep = ";")
#   
#   unlink(dir_out)
# }
```


```{r}
# save_data(n, y, dir_out = "C:/Users/maxs/Documents/data/aspol_scikit-learn/n")
# save_data(tf, y, dir_out = "C:/Users/maxs/Documents/data/aspol_scikit-learn/tf")
# save_data(tf_idf, y, dir_out = "C:/Users/maxs/Documents/data/aspol_scikit-learn/tf_idf")
```

