---
title: "2 Document term matrix"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Analysis expects input data to be in numerical form. Here we provide two examples 
of how to convert text data from tabular form to matrix format: one using simple 
bag-of-words approach and two ways using ngrams, namely bigram and trigram.

```{r setup, message=FALSE, warning=FALSE}
library(rfintext)
library(tidytext)
library(dplyr)
library(quanteda)
```

```{r, include=FALSE}
quanteda_options(print_dfm_max_ndoc = 20, print_dfm_max_nfeat = 20)
```

## Bag-of-Words representation

Pre processing `aspol` corpus and calculating document term count (*dtc*) using 
bag-of-words representation. Single term per document per row.

```{r, message=FALSE, R.options = list(width = 10000, pillar.print_max = 20, pillar.print_min = 20)}
bow <- aspol |>
  preprocess_corpus(kunta) |>
  count(kunta, LEMMA, name = "dtc")
bow
```

Converting original corpus to document term matrix.

```{r, message=FALSE, R.options = list(width = 10000)}
dtm <- bow |>
  cast_dfm(document = kunta, term = LEMMA, value = dtc)  # quanteda format
  # cast_dtm(document = kunta, term = LEMMA, value = dtc)  # tm format
dtm
```

It is possible to convert from one format to another with `quanteda::convert()` 
function. For example some functions require data to be in `tm` format.

```{r}
dtm |>
  quanteda::convert(to = "tm")
```

## Ngrams

Bigram and trigram

```{r, message=FALSE, R.options = list(width = 10000, pillar.print_max = 20, pillar.print_min = 20)}
ngram <- aspol |> 
  preprocess_corpus(kunta) |> 
  mutate(LEMMA2 = lead(LEMMA, default = "", n = 1), 
         LEMMA3 = lead(LEMMA, default = "", n = 2), 
         bigram = paste0(LEMMA, " ",  LEMMA2),
         trigram = paste0(LEMMA, " ", LEMMA2, " ", LEMMA3),
         .by = kunta)  # c(kunta, sent) limits the words to same sentence
count(ngram, bigram, sort = TRUE)
```

```{r, message=FALSE, R.options = list(width = 10000, pillar.print_max = 20, pillar.print_min = 20)}
count(ngram, trigram, sort = TRUE)
```

```{r}
ngram |>
  count(kunta, bigram, name = "n") |>
  cast_dfm(document = kunta, term = bigram, value = n)
```

```{r}
ngram |>
  count(kunta, trigram, name = "n") |>
  cast_dfm(document = kunta, term = trigram, value = n)
```

